---
title: "ROC and Reclassification analysis in R"
author: "Thor Aspelund, Icelandic Heart Association, University of Iceland Public Health"
date: "October 30th 2014"
output: beamer_presentation
subtitle: R Ísland meeting
---
```{r,echo=FALSE,message=FALSE}
library(xtable)
library(ggplot2)
library(epitools)
library(survival)
library(stargazer)
library(png)
library(rms)
library(Hmisc)
library(pROC)
library(PredictABEL)
library(survC1)
library(survIDINRI)
library(timeROC)

data(wcgs)
wcgs$smoker <- ifelse(wcgs$ncigs0>0,1,0)
wcgs$smokerfny <- factor(wcgs$smoker,levels=c(0,1),labels=c("No","Yes"))
wcgs$heightcm <- wcgs$height0*2.54
wcgs$weightkg <- wcgs$weight0*0.45359237
wcgs$bmi <- wcgs$weightkg / (wcgs$heightcm/100)^2
wcgs$cholmmol = wcgs$chol/39
wcgs$dibpat0f <- factor(wcgs$dibpat0,levels=0:1,label=c("B","A"))
wcgs$chd69fny <- factor(wcgs$chd69,levels=c(0,1),labels=c("No CHD","CHD"))
wcgs.s <- subset(wcgs,cholmmol<15)
```

```{r,cache=TRUE,echo=FALSE}
setwd("C:/Users/aspelund/Dropbox/Fyrirlestrar/R meeting ROC and Reclass 30okt14")
source("reclass.R")
```

## CHD risk models

* Icelandic Heart Association Risk Score
    + http://www.hjarta.is
  
* European HeartSCORE 
    + http://www.heartscore.org/Pages/welcome.aspx
  
* Framingham Risk Score (Circulation 1998) 
    + http://circ.ahajournals.org/content/97/18/1837.long

## The ongoing search for risk markers
* There is an existing risk score in use
* We would like to introcude a new marker to improve the score
* Is the new score better than the old score?  **That is the question**

## Introducing a new marker
### Measuring improvement using statistcal models
* Model 1: Basic risk score model 
* Model 2: Basic risk score model + new marker
* Is risk score 2 better than risk score 1?
* In other words: Is Model 2 an improvment of than Model 1?  

## How is a risk score evaluated?
* In categorical data analysis we have met various
concordance measures, such as:
    + Kendall's $tau_a$
    + Somer's D
    + The C index
  
##  Concordance measures - binary outcome
View $Y$ as a binary outcome, either $Y=1$ or $Y=0$.
Let $Z$ be a continous risk score.
The concordance between the risk score and the outcome can be
measured by Kendall's $\tau_a$
$$
\tau_a(Y,Z) = E(\text{sign}(Z_1-Z_2)\text{sign}(Y_1-Y_2)))
$$
where the pairs $(Y_1,Z_1)$ and $(Y_2,Z_2)$ are chosen at random. 

Somer's $D$ is an adaption of $\tau_a$ 
$$
D(Z,Y) = \tau_a(Y,Z) / \tau_a(Y,Y)
$$


## Concordance and the C statistic
If there are no ties in $Z$ it can be shown that
$$
D(Z,Y) = 2 \cdot P(Z_i > Z_j | Y_i > Y_j)-1 = 2 \cdot C(Z,Y) - 1
$$
Where 
$$
C(Z,Y) = P(Z_i > Z_j | Y_i > Y_j)
$$

* The $C$ statistic is the probability that a risk score for a case $Y_i=1$ is greater than the risk score for a control $Y_j=0$.  
* We want this probability to be high.  
* It can be shown (via integration) that $0.5<=C(Z,Y)<=1$.

## ROC analysis, AUC and the C statistic
* ROC analysis is an analysis of sensitivity and specificity
* Let $Z$ be a continuous risk score and $z$ be an arbitrary cutoff value. Choose a random case or a control (Denote by $Y$). 
  Compute the risk score $Z$ for each.
* Assume you are blinded to the case control status.  Declare the subject to be a case if Z>z, otherwise a control.
* Then $P(Z > z | Y=1)$ is the sensitivity or the True Positive Probability (TP).
* $P(Z > z | Y=0) = 1 - P(Z<z | Y=0)$ is 1-specificity or the False Positive Probability (FP).
* A graph of $TP=P(Z > z | Y=1)$ vs. $FP=P(Z > z | Y=0)$ is the ROC curve for the diagnostic test $Z$.
* It can be shown that the area under the curve (AUC) equals the $C$ statistic.  AUC=C!
* Common values for $C$ for risk models are in the range 0.70 to 0.75.

## Frank Harrell and the rms package
* Frank Harell provides many diagnostic functions in the *rms* package.  $C$ and $D$ are displayed as elements in 
*lrm* (logistic regression models) objects.  However, $C$ with $D$ and a standard error for $D$ are provided with *rcorr.cens*. Note that the standard error of $C$ is half the standard error of $D$.

* Recall:  $C=0.5 \cdot (D + 1)$

## The example data
* WCGS data - Western Collaborative Group Study
* Prospective study of heart disease among men in California, initiated in 1960 
* http://clinicaltrials.gov/show/NCT00005174
* N = 3154, age `r min(wcgs.s$age0)` to `r max(wcgs.s$age0)`, free of heart disease
* Follow-up for 10 years
* Data (*wcgs*) available via the *epitools* package
* N =  `r nrow(wcgs.s)` with complete data on risk factors used for analysis

## C and D results from the rms package
* Consider Model 0: chd ~ age
```{r,cache=TRUE}
fit0 <- lrm(chd69 ~ age0,data=wcgs.s,x=T,y=T)
data.frame(C=fit0$stats[6],D=fit0$stats[7])
rc0<-rcorr.cens(predict(fit0),fit0$y)
data.frame(Estimate=rc0[1],SE=rc0[3]/2,
           Lower=rc0[1]-1.96*rc0[3]/2,Upper=rc0[1]+1.96*rc0[3]/2)
```
We have $C=`r rc0[1]`$ for Model 0.  There is room for improvement.

## Are under the ROC curve for Model 0 and threshold
```{r,cache=TRUE}
roc0<-roc(fit0$y,predict(fit0),ci=T)
roc0.c <- coords(roc0,x="best",best.method=c("closest.topleft"))
roc0
roc0.c
```

## ROC curve for Model 0 and threshold
```{r,fig.height=4,fig.width=5,echo=FALSE,results='hide'}
plot(roc0,print.auc=T)
points(roc0.c[2],roc0.c[3],col="red",pch=16,cex=1.4)
```

## The threshold on the risk scale
```{r}
plogis(roc0.c[1])
```
The age where the threshold is reached
```{r}
data.frame(age=(roc0.c[1]-coef(fit0)[1])/coef(fit0)[2])
```

## Risk as a function of age
```{r,fig.height=4,fig.width=5}
plot(39:49,predict(fit0,newdata=data.frame(age0=39:49),type="fitted"),type="l",ylab="chd risk")
abline(v=(roc0.c[1]-coef(fit0)[1])/coef(fit0)[2])
text(45,0.05,paste("Vertical line at risk threshold, age =",(roc0.c[1]-coef(fit0)[1])/coef(fit0)[2]))
```

## Model 1 - add cholesterol, blood pressure, bmi, and smoking
```{r}
fit1<-update(fit0,.~.+cholmmol + sbp0 + bmi + smoker)
data.frame(C=fit1$stats[6],D=fit1$stats[7])
rc1<-rcorr.cens(predict(fit1),fit1$y)
data.frame(Estimate=rc1[1],SE=rc1[3]/2,
           Lower=rc1[1]-1.96*rc1[3]/2,Upper=rc1[1]+1.96*rc1[3]/2)
```

## Model 1 is an improvment
```{r,cache=TRUE,echo=FALSE}
roc1 <-roc(fit1$y,predict(fit1),ci=T)
roc1.c <- coords(roc1,x="best",best.method=c("closest.topleft"))
roc1
roc1.c
plogis(roc1.c[1])
```

## ROC curve for Model 1 and Model 0 and threshold
```{r,fig.height=4,fig.width=5,echo=FALSE,results='hide'}
plot(roc0,print.auc=F)
plot(roc1,add=T)
points(roc0.c[2],roc0.c[3],col="red",pch=16,cex=1.4)
points(roc1.c[2],roc1.c[3],col="blue",pch=16,cex=1.4)
```

## Test of improvement
```{r}
roc.test(roc0,roc1)
```

## The hunt for a new marker
* We have our basic risk model (Model 1)
* The $C$ statistic is `r rc1[1]`
* This is a typical value for a chd risk model
* We would still like to improve it
* The hunt is on for a new marker
* We add the new marker to Model 1 and measure the improvement

## Our new marker - Personality A vs B
Personality is associated with CHD.  The OR > 2.
```{r}
oddsratio.wald(wcgs$dibpat0f,wcgs$chd69)
```

## Models 
* Model 0: 0 chd ~ age
* Model 1: chd ~ age + bmi + chol + systolic + smoker
* Model 2: chd ~ age + bmi + chol + systolic + smoker + personality

--> personality with 2 levels (A,B) is our *new* marker

## Model 2 - Add the personality marker
```{r}
fit2 <- update(fit1,.~.+dibpat0f)
```
The adjusted OR is
```{r}
data.frame(OR=exp(coef(fit2)[7]),Lower=exp(confint.default(fit2)[7,1]),Upper=exp(confint.default(fit2)[7,2]))
```

## ROC curves 0 1 and 2
```{r,echo=FALSE,results='hide'}
roc2 <-roc(fit2$y,predict(fit2),ci=T)
roc2.c <- coords(roc2,x="best",best.method=c("closest.topleft"))
plot(roc0,col="red",lty=3)
points(roc0.c[2],roc0.c[3],col="red",pch=16,cex=1.4)
plot(roc1,add=T,lty=2,col="blue")
points(roc1.c[2],roc1.c[3],col="blue",pch=16,cex=1.4)
plot(roc2,add=T,lty=1,col="black",lwd=3,print.auc=T)
points(roc2.c[2],roc2.c[3],col="green",pch=16,cex=1.4)
legend(0.4,0.4,lty=c(3,2,1),col=c("red","blue","black"),legend=c("Model 0","Model 1","Model 2"))
```

## Likelihood ratio test comparing Model 1 & 2
```{r}
lrtest(fit1,fit2)
```
* This means that the marker is highly significant.
* Recall that the $\text{OR} \approx 2$

## The Pepe 2004 paper
* Limitations of the Odds Ratio in Gauging the Performance of a Diagnostic, Prognostic, or Screening Marker
* Margaret Sullivan Pepe, Holly James, Gary Longton, Wendy Leisenring, and Polly Newcomb
* American Journal of Epidemiology 2004

### Message
* Tells us about the limitations of the OR as a measure of diagnostic capacity and that ROC curves and sensitivity and specificity must be studied.
* Also demonstrates how difficult it is to see a change in ROC curves between models using 1 new marker


## Formally comparing ROC curves 1 and 2
```{r}
roc.test(roc1,roc2)
```
* Statisticsally signficant
* The increment is less than 0.02!

## The NEJM paper from the Icelandic Heart Association 
* C-reactive protein and other circulating markers of inflammation in the prediction of coronary heart disease.
* Danesh J1, Wheeler JG, Hirschfield GM, Eda S, Eiriksdottir G, Rumley A, Lowe GD, Pepys MB, Gudnason V.
* N Engl J Med. 2004 Apr 1;350(14):1387-97.

### Message
* CRP is a statistically significant marker
* Adding CRP to a risk model using traditional risk factors increase the ROC area by 0.01
* *C-reactive protein is a relatively moderate predictor of coronary heart disease. Recommendations regarding its use in predicting the likelihood of coronary heart disease may need to be reviewed*

## Frustration among CRP advocates
What are we going to do about these small increments in AUC?

## Nancy R Cook paper 2006 - Reclassification
* The effect of including C-reactive protein in cardiovascular risk prediction models for women.
* Cook NR1, Buring JE, Ridker PM.
* Ann Intern Med. 2006 Jul 4;145(1):21-9.

### Message
* Introduced the concept of **reclassification**
* Do subjects move between risk categories after adding the predictor?
* A global risk prediction model that includes hsCRP improves cardiovascular risk classification in women, particularly among those with a 10-year risk of 5% to 20%. In models that include age, blood pressure, and smoking status, hsCRP improves prediction at least as much as do lipid measures.

## Nancy R Cook paper 2006 - Reclassification
* Did I mention the conflict of interest?

### Potential conflict of interest reported in Cook's paper
Dr. Ridker is listed as a co-inventor on patents held by the Brigham and Women's Hospital that relate to the use of inflammatory biomarkers in cardiovascular disease.

```{r,results='hide',echo=FALSE}
pred1 <- predict(fit1,type="fitted")
pred2 <- predict(fit2,type="fitted")

pred1c <- cut(pred1*100,c(0,10,20,100))
pred2c <- cut(pred2*100,c(0,10,20,100))

table(pred1c)
table(pred2c)
tblc <- table(pred1c,pred2c)
```

## Example of Cook's approach
```{r}
tblc
```
* This shows that many participants are reclassified. 
* For example: `r tblc[1,2]` are reclassfied from 0 to 10% risk into 10-20% risk


## Peninca 2008
* Have to consider reclassification of people who develop and who do not develop the events **separately**
* Defines the net reclassification improvement NRI based on risk categories
```{r,results='hide',echo=FALSE}
tblcs <- table(pred1c,pred2c,wcgs.s$chd69)
tblcsc <- cbind(tblcs[,,1],tblcs[,,2])
StatusCHD <- c("","CHD=0","","","CHD=1","")
tblcscr<-rbind(StatusCHD,tblcsc)
```

```{r,results='asis',message=FALSE,echo=FALSE}
print(xtable(tblcscr),type="latex",comment=FALSE)
```

* Inroduces statistical inference about reclassification (NRI)
and Integrated discrimination improvement (IDI)


## Test of net reclassification NRI
Asymptotic test of


$NRI = (\hat{p}_{up,events}-\hat{p}_{down,events})-(\hat{p}_{up,nonevents}-\hat{p}_{down,nonevents})$

* Notice the retrospective definition
* Doesn't really apply to **case control** data unless we can adjust the risk estimates to be meaningful

## NRI estimate
* Using *reclass* from
* http://www.ucr.uu.se/en/index.php/epistat/program-code/306-nri-and-idi

```{r}
rcls<-reclass(chd69 ~ age0 + cholmmol + sbp0 + bmi + smoker 
              + dibpat0f,lim=c(0.1,0.2),wcgs.s,1,TRUE)
```

the estimate was `r rcls$NRI` with $95\%$ CI as 
(`r rcls$NRI-1.96*rcls$SE.NRI`,`r rcls$NRI+1.96*rcls$SE.NRI`).  In fact 
`r rcls$reclass.without.outcome` in without event and `r rcls$reclass.outcome` in with outcome.

## Test of IDI
Asymptotic test of the difference in difference between risk of non-cases and cases

$IDI = (\bar{\hat{p}}_{new,events}-\bar{\hat{p}}_{new,nonevents})-(\bar{\hat{p}}_{down,events}-\bar{\hat{p}}_{down,events})$.  Estimate = `r rcls$IDI` with SE = `r rcls$SE.IDI`.

```{r,echo=FALSE,fig.height=3,fig.width=5}
idiframe <- data.frame(model=c(rep("M1",length(pred1)),rep("M2",length(pred2))),
                       risk=c(pred1,pred2),status=c(wcgs.s$chd69f,wcgs.s$chd69f))

p <- ggplot(idiframe, aes(factor(status,labels=c("No CHD","CHD")), risk)) + xlab("")
p <- p + theme(text = element_text(size=14))
p + geom_boxplot(aes(fill = factor(model,labels=c("Model 1","Model 2")))) +  scale_fill_discrete(name="Model")
```


## Adding a new marker - Statistics to report
* JAMA 2009 Review paper:
* Assessment of Claims of Improved Prediction Beyond the Framingham Risk Score
* Ioanna Tzoulaki, PhD George Liberopoulos, MD John P. A. Ioannidis, MD

  -> Set standard

* Akaike Information Criteria (AIC)
* AUCs with and without the new predictor
* Difference in AUC with a confidence interval
* Calibration with and without the new marker with a goodness of fit test
* Documentation of **reclassification**

