---
title: "ROC and Reclassification analysis in R - part 2"
author: "Thor Aspelund, Icelandic Heart Association, University of Iceland Public Health"
date: "October 30th 2014"
output: beamer_presentation
subtitle: R Ísland meeting
---
```{r,echo=FALSE,message=FALSE}
library(xtable)
library(ggplot2)
library(epitools)
library(survival)
library(stargazer)
library(png)
library(rms)
library(Hmisc)
library(pROC)
library(PredictABEL)
library(survC1)
library(survIDINRI)
library(timeROC)

data(wcgs)
wcgs$smoker <- ifelse(wcgs$ncigs0>0,1,0)
wcgs$smokerfny <- factor(wcgs$smoker,levels=c(0,1),labels=c("No","Yes"))
wcgs$heightcm <- wcgs$height0*2.54
wcgs$weightkg <- wcgs$weight0*0.45359237
wcgs$bmi <- wcgs$weightkg / (wcgs$heightcm/100)^2
wcgs$cholmmol = wcgs$chol/39
wcgs$dibpat0f <- factor(wcgs$dibpat0,levels=0:1,label=c("B","A"))
wcgs$chd69fny <- factor(wcgs$chd69,levels=c(0,1),labels=c("No CHD","CHD"))
wcgs.s <- subset(wcgs,cholmmol<15)
```

```{r,cache=TRUE,echo=FALSE}
setwd("C:/Users/aspelund/Dropbox/Fyrirlestrar/R meeting ROC and Reclass 30okt14")
source("reclass.R")
```

```{r,echo=FALSE,results='hide'}
rcls<-reclass(chd69 ~ age0 + cholmmol + sbp0 + bmi + smoker 
              + dibpat0f,lim=c(0.1,0.2),wcgs.s,1,TRUE)
```

## The category free NRI (continuous NRI)
* Statistics in Medince 2011 (online 2010)
* Extensions of net reclassification improvement
calculations to measure usefulness of new
biomarkers
* Michael J. Pencina, Ralph B. D Agostino Sr, and Ewout W. Steyerberg

$$
NRI= \frac{(P(event|up)−P(event))P(up)+(P(event)−P(event|down))P(down)}{P(event)(1−P(event))} 
$$

* Notice the prospective defintion
* It can be interpreted as a measure of
event rate increase among those who are reclassified upwards and event rate decrease among those who are reclassified
downwards
* Applies to survival and case control data

```{r,cache=TRUE,echo=FALSE,results='hide'}
fit0 <- lrm(chd69 ~ age0,data=wcgs.s,x=T,y=T)
fit1<-update(fit0,.~.+cholmmol + sbp0 + bmi + smoker)
fit2 <- update(fit1,.~.+dibpat0f)
pred1 <- predict(fit1,type="fitted")
pred2 <- predict(fit2,type="fitted")
```


## The continuous NRI
* From PredictABEL
* reclassification(wcgs.s,11,pred1,pred2,c(0,0.1,0.2,1))
* NRI(Continuous) [95% CI]: 0.4081 [ 0.2892 - 0.527 ] ; p-value: 0


## Survival data
```{r}
cox1 <- cph(Surv(time169,chd69) ~ age0 + cholmmol + sbp0 + smokerfny,data=wcgs.s,x=T,y=T)
cox2 <- cph(Surv(time169,chd69) ~ age0 + cholmmol + sbp0 + smokerfny + dibpat0f,data=wcgs.s,x=T,y=T)
lrtest(cox1,cox2)
```

## The C statistic comparison (Harrell)
```{r}
rcorrp.cens(1-predict(cox1,type="lp"),1-predict(cox2,type="lp"),Surv(wcgs.s$time169,wcgs.s$chd69))
```

## The C statistics (survC1)
```{r,cache=TRUE}
mydata1<-cbind(wcgs.s$time169,wcgs.s$chd69,cox1$x)
mydata2<-cbind(wcgs.s$time169,wcgs.s$chd69,cox2$x)
sc1<-Est.Cval(mydata1,5*365.24)
sc2<-Est.Cval(mydata2,5*365.24)
names(sc1)
sc1$Dhat
sc2$Dhat
```

## C statistics  (timeROC)
```{r,cache=TRUE}
troc1<-timeROC(wcgs.s$time169,wcgs.s$chd69,predict(cox1,type="lp"),cause=1,times=5*365.24,iid=T)
troc2<-timeROC(wcgs.s$time169,wcgs.s$chd69,predict(cox2,type="lp"),cause=1,times=5*365.24,iid=T)
troc1
troc2
```

## Comparisons (timeROC)
```{r,cache=TRUE}
compare(troc1,troc2)
```


## Pepe paper 2013
* Statistics in Medicine 2013
* Pepe MS 
* Tests of significance of parameter, change in AUC, NRI, IDI all equivalent
* Makes no sense to test the same hypothesis many times
* Report AUC and change as descriptive statistics and focus on practical change in magnitude
* Show reclassification tables rather than the continuous NRI

## Conclusion
* Is Jörgen Hilden right about these being good looking statistics with nothing underneath?
* Yes, because all the statistical signficance to test is for the parameter
* Don't dwell on the statistical significance of difference in AUC, NRI and IDI
* The additional statistics are still good descriptive statistics of model performance
* The AUC is a fundamental statistic to report and the inteplay between sensitivity and specificity
* Decision theory (papers by Vickers and Steyerberg)


